---
title: "Analysis of bulk RNA-seq data: Signatures of Fibrosis"
format:
  html:
    theme: cosmo
    toc: true
    df-print: kable
date: today
---

# Before we start: check your setup


```{r}
#| message: false
pkgs <- c("tidyverse", "ggpubr", "here", "data.table", "DESeq2", "matrixStats",
          "limma", "clusterProfiler", "org.Hs.eg.db", "ggrepel")

missing_pkgs <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install(missing_pkgs, update = FALSE, ask = FALSE)
}

library(here)
need_files <- c("bulk/cdna/t2g.csv", "bulk/GSE135251_samples.csv", "bulk/GSE135251")
missing_files <- need_files[!file.exists(here(need_files))]
if (length(missing_files))
  stop("Missing data (check you opened the project's .Rproj): ",
       paste(missing_files, collapse = ", "))

message("Setup OK. here() root = ", here())
```

# Generating kallisto index

```{r, eval=FALSE}
# run in terminal
# mamba install kallisto
# ./kallisto/kallisto index -t 8 -i human_cdna.idx cdna/Homo_sapiens.GRCh38.cdna.all.fa.gz
kallisto index -t 8 -i human_cdna.idx cdna/Homo_sapiens.GRCh38.cdna.all.fa.gz
```

```{r, eval=FALSE}
# run in terminal

kallisto quant -i human_cdna.idx -o rnaseq_data/kallisto_output/ -t 12 rnaseq_data/fastq/SRR9883171_1.fastq.gz rnaseq_data/fastq/SRR9883171_2.fastq.gz
```

```{r, eval=FALSE}
# run in terminal
cd output
ls -ltrha
```

```{r, eval=FALSE}
# run in terminal
head abundance.tsv
```

```{r, eval=FALSE}
# run in terminal
 zcat scripts/Homo_sapiens.GRCh38.cdna.all.fa.gz| head
```

```{r, eval=FALSE}
# run in terminal
cd cdna
zcat Homo_sapiens.GRCh38.cdna.all.fa.gz| \
grep '>' |
awk '{FS= " "}BEGIN{ print "tx_id,gene_id,gene_name"};{print substr($1,2) "," substr($4,6) "," substr($7,13)};' > t2g.csv
```

You can also download the `t2g.csv` from here: https://www.dropbox.com/scl/fi/dpbslasta98e05tru1ao6/t2g.csv?rlkey=xp10abwca1om9zn5n8frnogn5&st=qxkb672p&dl=0

```{r, eval=FALSE}
wget --content-disposition -c "https://www.dropbox.com/scl/fi/dpbslasta98e05tru1ao6/t2g.csv?rlkey=xp10abwca1om9zn5n8frnogn5&st=qxkb672p&dl=0"
```

Download the metadata from  https://seqout.org/p/GSE135251

Make sure you have 216 samples in the dataset

Also available [here](https://www.dropbox.com/scl/fi/uk6o1s2cegl77gcma7r81/GSE135251_samples.csv?rlkey=2nzbodzezbpx5qm79t35stwr8&st=m801enqd&dl=0)



# Analysis of kallisto counts

```{r}
library(tidyverse)
library(ggpubr)
library(here)

theme_set(theme_pubr())
# change file paths as required

geo.counts <- read_tsv(here("bulk/GSE135251/GSM3998232_17_82.counts.txt.gz"), col_names = c("gene_id", "geo_counts")) 
head(geo.counts)
```

```{r}
# change file paths as required

t2g <- read_csv(here("bulk/cdna/t2g.csv"))
head(t2g)
```

```{r}
# change file paths as required
kallisto.tx.counts <- read_tsv(here("bulk/rnaseq_data/kallisto_output/abundance.tsv"))
head(kallisto.tx.counts)
```

```{r}
# change file paths as required
kallisto.gene.counts <- kallisto.tx.counts %>%
    inner_join(t2g, by = c("target_id" = "tx_id")) %>% 
    mutate(gene_id = str_replace(gene_id, "\\..*", "")) %>% # remove .1, .2 etc
    group_by(gene_id, gene_name) %>% # group by gene id
    summarise(kallisto_counts = sum(est_counts)) %>% # sum the counts across all isoforms of a gene
    ungroup() 

kallisto.geo.counts <- full_join(kallisto.gene.counts, geo.counts)
```

```{r}
library(ggpubr)
ggplot(kallisto.geo.counts, aes(x = kallisto_counts, y = geo_counts)) +
    geom_point() +
    geom_abline() +
    labs(x = "Kallisto counts", y = "GEO counts") +
    scale_x_log10() +
    scale_y_log10() +
    theme_pubr()
```

```{r}
kallisto.geo.counts.inner <- inner_join(kallisto.gene.counts, geo.counts)
cor(kallisto.geo.counts.inner$kallisto_counts, kallisto.geo.counts.inner$geo_counts)
```

# Differential expression analysis

```{bash, eval=FALSE}
cd GSE135251
ls -1 | wc -l
```

```{r}
library(data.table)

# Read every per-sample file. Each is 2 columns, no header: Ensembl gene_id + count.
files <- list.files(here("bulk/GSE135251/"),
                    pattern = "\\.counts\\.txt\\.gz$", full.names = TRUE)
stopifnot(length(files) > 0)
z <- lapply(X = files, FUN = fread, header = FALSE,
            col.names = c("gene_id", "count"))

# Name each element by its GSM accession (first "_"-delimited token of the filename)
names(z) <- sub("_.*$", "", basename(files))
stopifnot(!anyDuplicated(names(z)))
```

```{r}
# Long -> wide: one row per gene_id, one column per sample (GSM), values = counts.
counts.by.id <- rbindlist(z, idcol = "sample") %>%
    filter(!str_starts(gene_id, "__")) %>%                    # drop HTSeq summary rows
    mutate(gene_id = str_replace(gene_id, "\\..*", "")) %>%   # drop .1/.2 version suffix
    group_by(gene_id, sample) %>%
    summarise(count = sum(count), .groups = "drop") %>%       # collapse any duplicate ids
    pivot_wider(names_from = sample, values_from = count, values_fill = 0)
```

```{r}
# Map Ensembl gene_id -> gene symbol.
suppressPackageStartupMessages(library(org.Hs.eg.db))

symbol_map <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys      = counts.by.id$gene_id,
  keytype   = "ENSEMBL",
  column    = "SYMBOL",
  multiVals = "first"
)

all_counts <- counts.by.id %>%
    mutate(gene_name = unname(symbol_map[gene_id]),
           gene_name = if_else(is.na(gene_name) | gene_name == "",
                               gene_id, gene_name)) %>%        # no symbol -> use the Ensembl id
    dplyr::select(gene_name, everything(), -gene_id) %>%    
    dplyr::relocate(gene_name)

head(all_counts)
```

```{r}
counts <- all_counts %>% as.data.frame()
head(counts)
```

Altenatively, you can also download counts from [here](https://www.dropbox.com/scl/fi/om24ym8s2ehw0uqn5jpdf/all_counts.zip?rlkey=eh6a1jy35t33zro1dlwrp9w12&st=zv7j6iye&dl=0)

```{r}
rownames(counts) <- make.unique(counts$gene_name)
counts$gene_name <- NULL
```

```{r}
metadata <- read_csv(here("bulk/GSE135251_samples.csv")) %>% as.data.frame()
head(metadata)
```

```{r}
ncol(metadata)
```

```{r}
table(metadata$Stage)
```

```{r}
rownames(metadata) <- metadata$Sample

# keep only samples we have both counts AND metadata for
shared <- intersect(colnames(counts), metadata$Sample)
message(length(colnames(counts)), " count samples, ",
        nrow(metadata), " annotated samples, ",
        length(shared), " used (present in both).")
stopifnot(length(shared) > 0)

counts   <- counts[, shared, drop = FALSE]
metadata <- metadata[shared, ]                       # reorder to match counts
metadata$sample_name <- rownames(metadata)

stopifnot(identical(rownames(metadata), colnames(counts)))

# design variable: Stage has control / early / moderate; control is the reference
metadata$stage <- relevel(factor(metadata$Stage), ref = "control")
table(metadata$stage)
```

```{r}
library(DESeq2)
# DESeq2 needs an integer count matrix; kallisto est_counts can be fractional
count_mat <- round(as.matrix(counts))
storage.mode(count_mat) <- "integer"

dds <- DESeqDataSetFromMatrix(countData = count_mat,
                              colData = metadata,
                              design= ~ stage)
smallestGroupSize <- 3
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]
```

```{r}
dds <- DESeq(dds)
```

## Exploratory analysis

```{r}
vsd <- vst(dds, blind=FALSE)
```

Can we normalise the data by just taking log?

```{r}
df <- data.frame(mean = log1p(rowMeans(counts)),
variance = matrixStats::rowVars(log1p( counts %>% as.matrix())))

ggplot(df, aes(mean, variance)) + geom_point() + scale_x_log10() + scale_y_log10()
```

How does DESeq2 achieve variance stabilisation?

```{r}
df <- data.frame(mean = rowMeans(counts), variance = matrixStats::rowVars(counts %>% as.matrix()))

ggplot(df, aes(mean, variance)) + geom_point() + scale_x_log10() + scale_y_log10()
```

```{r}
df.vsd <- data.frame(mean = rowMeans(assay(vsd)),
variance = matrixStats::rowVars(assay(vsd) %>% as.matrix()))

ggplot(df.vsd, aes(mean, variance)) + geom_point() + scale_x_log10() + scale_y_log10()
```


```{r}
vsd$group.in.paper <- vsd$`group in paper`
plotPCA(vsd, intgroup=c('stage'))
```

```{r}
plotPCA(vsd, intgroup=c("stage"))
```

### PCA by hand


`plotPCA` keeps the top 500 most variable genes, so we do the same, center each
gene (rows), and run `prcomp` on the transposed matrix (samples as rows).

```{r}
mat <- assay(vsd)
top_genes <- head(order(matrixStats::rowVars(mat), decreasing = TRUE), 500)
mat_top <- mat[top_genes, ]

pca <- prcomp(t(mat_top), center = TRUE, scale. = FALSE)

pct <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)
pct[1:5]
```

`prcomp` itself is just an SVD of the centered matrix, $X = U D V^{T}$:

```{r}
X  <- scale(t(mat_top), center = TRUE, scale = FALSE)  # samples x genes, gene-centered
sv <- svd(X)

# pca$x  (sample coordinates)
scores_svd <- sv$u %*% diag(sv$d)
# pca$rotation (gene loadings)
load_svd   <- sv$v   
# pca$sdev
sdev_svd   <- sv$d / sqrt(nrow(X) - 1) 

# same up to an arbitrary per-PC sign flip
all.equal(abs(scores_svd[, 1:5]), abs(unname(pca$x[, 1:5])))
all.equal(sdev_svd, pca$sdev)
round(100 * sdev_svd^2 / sum(sdev_svd^2), 1)[1:5]
```

Plot PC1 vs PC2:

```{r}
scores <- as.data.frame(pca$x)
scores$stage <- vsd$stage

ggplot(scores, aes(PC1, PC2, color = stage)) +
  geom_point(size = 2) +
  labs(x = paste0("PC1 (", pct[1], "%)"),
       y = paste0("PC2 (", pct[2], "%)")) +
  theme_pubr()
```

### Visualising the loadings

The **loadings** (`pca$rotation`) give each gene's contribution to a PC. Genes with the largest |loading| define that PC.

```{r}
loadings <- as.data.frame(pca$rotation)
loadings$gene <- rownames(loadings)

# top genes driving PC1 and PC2
top_pc1 <- loadings %>% arrange(desc(abs(PC1))) %>% head(15)
top_pc2 <- loadings %>% arrange(desc(abs(PC2))) %>% head(15)
top_pc1$gene
top_pc2$gene
```

```{r}
ggplot(top_pc2, aes(x = reorder(gene, PC2), y = PC2)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top genes loading on PC2", x = NULL, y = "PC2 loading") +
  theme_pubr()
```

A loadings biplot: samples as points, with the strongest PC2 gene arrows overlaid.

```{r}
arrows <- top_pc2 %>% arrange(desc(abs(PC2))) %>% head(8)
scale_f <- 0.9 * max(abs(scores$PC2)) / max(abs(arrows$PC2))   # scale arrows to score range

ggplot(scores, aes(PC1, PC2)) +
  geom_point(aes(color = stage), size = 2) +
  geom_segment(data = arrows,
               aes(x = 0, y = 0, xend = PC1 * scale_f, yend = PC2 * scale_f),
               arrow = arrow(length = unit(0.15, "cm")), color = "grey30") +
  ggrepel::geom_text_repel(data = arrows,
               aes(x = PC1 * scale_f, y = PC2 * scale_f, label = gene), size = 3) +
  labs(x = paste0("PC1 (", pct[1], "%)"), y = paste0("PC2 (", pct[2], "%)")) +
  theme_pubr()
```

Look up some gene's expression

```{r}
df = plotCounts(dds, gene="XIST", intgroup="sample_name", returnData = T)
df
```

```{r}
hist(df$count)
```

```{r}
library(ggplot2)
ggplot(df, aes(count)) + stat_ecdf(geom = "point") + theme_classic(base_size = 14)
```

```{r}
male_samples <- df[df$count<20,]$sample_name
```

```{r}
vsd$sex <- "female"
vsd$sex[vsd$sample_name %in% male_samples] <- "male"
```

```{r}
plotPCA(vsd, intgroup=c("sex"))
```

```{r}
plotPCA(vsd, intgroup=c("disease"))
```

```{r}
vsd.fixed <- vsd
assay(vsd.fixed) <- limma::removeBatchEffect(assay(vsd), batch = vsd$sex)
plotPCA(vsd.fixed, intgroup=c("stage"))
```

## DE analysis

```{r}
resultsNames(dds)   
```

```{r}
# reference is "control"; test a disease stage against it
case_stage <- "moderate"
de_results <- results(dds, contrast = c("stage", case_stage, "control"))
de_results <- de_results %>% as.data.frame()
```

```{r}
de_results$gene <- rownames(de_results)
de_results.pos <- de_results %>% as.data.frame() %>% dplyr::filter(log2FoldChange > 1) %>% dplyr::filter(padj < 0.05)
```

```{r}
library(clusterProfiler)
library(org.Hs.eg.db) # install using BiocManager::install("org.Hs.eg.db")
de.genes <- head(de_results.pos$gene, 500)   
universe.genes <- rownames(counts)
ego <- enrichGO(gene          = de.genes,
                universe      = universe.genes,
                OrgDb         = org.Hs.eg.db, 
                keyType = "SYMBOL",
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
        readable      = TRUE)
head(ego)

```

```{r}
barplot(ego)
```

# Volcano Plots

```{r}
# draw volcano plot from deseq2 results
library(ggplot2)
library(ggrepel)
de_results$gene <- rownames(de_results)
de_results$gene_name <- ifelse(de_results$padj < 0.05 & abs(de_results$log2FoldChange) > 1, de_results$gene, "")
de_results$group <- ifelse(de_results$padj < 0.05 & abs(de_results$log2FoldChange) > 1, "DE", "not-DE")
de_results$size <- ifelse(de_results$padj < 0.05 & abs(de_results$log2FoldChange) > 1, 3, 1)
ggplot(de_results, aes(x = log2FoldChange, y = -log10(padj),
                       color = group)) +
    geom_point(size=0.1) +
    geom_text_repel(aes(label = gene_name), box.padding = 0.5, show.legend = F) +
    scale_color_brewer(palette = "Set1") +
    labs(x = "log2 Fold Change", y = "-log10(padj)") +
    theme_minimal()

```

