---
title: "Multiple hypothesis testing: A primer"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
---

## The problem of multiple testing

Say you are developing a diagnostic panel and have 500 candidate genes that might be associated with cancer risk. You want to know which genes are truly associated and which associations arose by chance.

For each gene you measure expression in cases vs. controls and run a statistical test (say a t-test).

One test at α = 0.05 carries a 5% chance of a false positive (Type I error). With 500 independent tests at α = 0.05:

$$\Pr(\text{at least one false positive}) = 1 - (1 - 0.05)^{500} = 1 - (0.95)^{500} \approx 0.99999$$

So you will end up something significant (by chance) even when nothing really is.

## Why does this happen?

Think of each null gene test as a biased coin that lands heads (a false positive) with probability α = 0.05. One flip rarely comes up heads. But  if you flip 500 such coins once each you expect $500 \times 0.05 = 25$ heads. So what was a rare head per-coin event becomes common once you have enough coins.

So if all 500 genes were truly null, you would still expect ~25 to cross p < 0.05 by chance.

Now suppose the panel flags 30 genes and only a handful of the 500 are genuinely associated. About 25 of the 30 are chance hits and only ~5 are real, an 83% false discovery rate. Reporting all 30 would give patients misleading results and send everyone chasing false leads.

This is the **multiple testing problem**.

```{r setup}
library(ggplot2)
library(dplyr)
library(tidyr)

# Set seed for reproducibility
set.seed(42)

mytheme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.title = element_text(size = 11),
    legend.position = "top"
  )
```

### Visualising the problem

Simulate 100 tests where every null is true:

```{r multiple-testing-demo}
n_tests <- 100
p_values <- runif(n_tests, 0, 1) # Under null, p-values are uniformly distributed

# Count "significant" results at alpha = 0.05
n_significant <- sum(p_values < 0.05)

# Create histogram
hist_data <- data.frame(p_value = p_values)

ggplot(hist_data, aes(x = p_value)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "black") +
  geom_vline(xintercept = 0.05, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text",
    x = 0.05, y = Inf,
    label = paste0(n_significant, " 'significant' \nfalse positives"),
    hjust = -0.1, vjust = 1.5, color = "red", size = 4
  ) +
  labs(
    title = "P-values from 100 Tests Where Null is True",
    subtitle = "Red line shows α = 0.05 threshold",
    x = "P-value",
    y = "Count"
  ) +
  mytheme
```

We found `{r} n_significant` "significant" results with nothing real behind them.

## Types of error

Across many tests, four outcomes are possible:

```{r error-types-table}
error_matrix <- data.frame(
  Reality = c(
    "Null True (No Effect)", "Null True (No Effect)",
    "Null False (Real Effect)", "Null False (Real Effect)"
  ),
  Decision = c("Not Reject", "Reject", "Not Reject", "Reject"),
  Outcome = c(
    "Correct ✓", "FALSE POSITIVE ✗\n(Type I Error)",
    "FALSE NEGATIVE ✗\n(Type II Error)", "Correct ✓"
  ),
  Color = c("good", "bad", "bad", "good")
)

knitr::kable(error_matrix[, 1:3],
  col.names = c("Reality", "Your Decision", "Outcome"),
  caption = "Possible Outcomes in Hypothesis Testing"
)
```

### Two error rates

**Family-Wise Error Rate (FWER)** is the probability of making even one false positive across the whole family of tests. Controlling it is conservative: you make few false calls but miss real effects. Bonferroni controls FWER.

**False Discovery Rate (FDR)** is the expected fraction of your discoveries that are false. Controlling it is more permissive: you tolerate, say, 5% false calls among the ones you report, and in exchange you recover more real effects. Benjamini-Hochberg controls FDR.

### Clinical setting

## The Benjamini-Hochberg procedure

### The algorithm

1.  Sort all p-values from smallest to largest
2.  Number them: i = 1, 2, 3, ..., m (where m is total number of tests)
3.  Calculate threshold for each test: (i/m) × α (where α is your desired FDR, usually 0.05)
4.  Find the largest i where: p-value ≤ (i/m) × α
5.  Reject all hypotheses from 1 to i

A worked simulation:

```{r bh-visual-demo}
#  100 tests, 20 are truly significant
set.seed(123)
n_tests <- 100
n_true <- 20
n_null <- n_tests - n_true

# Generate p-values
# True alternatives: small p-values (from exponential distribution)
p_true <- rbeta(n_true, 0.5, 10)
# True nulls: uniform p-values
p_null <- runif(n_null, 0, 1)

# Combine and create dataset
p_values <- c(p_true, p_null)
true_status <- c(rep("True Effect", n_true), rep("No Effect", n_null))

test_data <- data.frame(
  test_id = 1:n_tests,
  p_value = p_values,
  truth = true_status
) %>%
  arrange(p_value) %>%
  mutate(
    rank = row_number(),
    bh_threshold = (rank / n_tests) * 0.05,
    bonf_threshold = 0.05 / n_tests,
    significant_bh = p_value <= bh_threshold,
    significant_bonf = p_value <= bonf_threshold
  )

# Find the BH cutoff
bh_cutoff <- max(which(test_data$significant_bh))

# Create the visualization
ggplot(test_data, aes(x = rank, y = p_value)) +
  # BH threshold line
  geom_line(aes(y = bh_threshold), color = "blue", linewidth = 1, linetype = "solid") +
  # Bonferroni threshold line
  geom_hline(yintercept = 0.05 / n_tests, color = "orange", linewidth = 1, linetype = "dashed") +
  # Alpha line
  geom_hline(yintercept = 0.05, color = "gray", linewidth = 0.5, linetype = "dotted") +
  # Points colored by truth
  geom_point(aes(color = truth, shape = truth), size = 3, alpha = 0.7) +
  # Highlight the BH cutoff
  geom_vline(xintercept = bh_cutoff, color = "blue", linetype = "dotted", alpha = 0.5) +
  # Annotations
  annotate("text",
    x = 80, y = 0.045, label = "BH Threshold\n(i/m) × α",
    color = "blue", size = 4, fontface = "bold"
  ) +
  annotate("text",
    x = 80, y = 0.0005, label = "Bonferroni\nα/m",
    color = "orange", size = 4, fontface = "bold"
  ) +
  annotate("text",
    x = bh_cutoff, y = 0.8,
    label = paste0("BH rejects ", bh_cutoff, " tests"),
    color = "blue", hjust = -0.1, size = 3.5
  ) +
  scale_color_manual(values = c("True Effect" = "#E74C3C", "No Effect" = "#95A5A6")) +
  scale_shape_manual(values = c("True Effect" = 16, "No Effect" = 1)) +
  labs(
    title = "Benjamini-Hochberg Procedure Visualization",
    subtitle = "Blue line shows BH threshold increasing with rank",
    x = "Rank (i) - sorted from smallest to largest p-value",
    y = "P-value",
    color = "True Status",
    shape = "True Status"
  ) +
  mytheme +
  theme(legend.position = "right")
```

### Reading the plot

The blue line is the BH threshold, which rises with rank:

-   On the left, p-values are small and easily fall under the (low) threshold.
-   On the right, the threshold is higher, so it can reject larger p-values.
-   BH finds the largest rank where the p-value is still below the line, and rejects everything up to it.

```{r bh-results-table}
results_summary <- test_data %>%
  group_by(truth) %>%
  summarise(
    Total = n(),
    `BH Significant` = sum(significant_bh),
    `Bonf Significant` = sum(significant_bonf)
  )

knitr::kable(results_summary,
  caption = "Comparison: True effects vs. Detected effects"
)
```

BH found `{r} sum(test_data$significant_bh)` significant tests; Bonferroni found `{r} sum(test_data$significant_bonf)`; the truth is `{r} n_true` real effects. BH recovers more of the real effects while still controlling false discoveries.

## Why does this work?

### The ranking creates protection

True effects tend to have small p-values, so they cluster at the low ranks, while a true null gives a Uniform(0,1) p-value that could land anywhere. Sorting groups most real effects near the front. The threshold at rank $i$ is $(i/m)\alpha$, rising by only $\alpha/m$ per step. Because null p-values are uniform, the chance that any one null falls under the threshold is just the threshold itself, so across all $m_0$ nulls the expected number of false rejections stays bounded by $m_0\alpha/m \le \alpha$ times the number of discoveries. That is what caps the FDR while the early-ranking real effects still get rejected.

### The stopping rule

```{r staircase-visual}
# Zoom in on the first 30 tests
zoom_data <- test_data %>% filter(rank <= 30)

ggplot(zoom_data, aes(x = rank, y = p_value)) +
  # Step function showing BH decision
  geom_step(aes(y = bh_threshold), color = "blue", linewidth = 1.5, direction = "mid") +
  geom_point(aes(color = truth, shape = significant_bh), size = 4, alpha = 0.8) +
  geom_segment(aes(x = rank, xend = rank, y = 0, yend = p_value, color = truth),
    alpha = 0.3, linetype = "dotted"
  ) +
  scale_color_manual(values = c("True Effect" = "#E74C3C", "No Effect" = "#95A5A6")) +
  scale_shape_manual(
    values = c("TRUE" = 19, "FALSE" = 1),
    guide = guide_legend(title = "BH Rejects?")
  ) +
  labs(
    title = "Close-up: How BH makes decisions",
    subtitle = "Points below the blue line are rejected (significant)",
    x = "Rank (i)",
    y = "P-value",
    color = "True Status"
  ) +
  mytheme +
  theme(legend.position = "right")
```

BH finds the *largest* rank whose p-value still sits under the threshold, then rejects every test up to that rank.

## The mathematical intuition

### Why the (i/m) × α formula?

The threshold is $p_{(i)} \le (i/m)\,\alpha$, where:

-   $i$ = the rank of this p-value (1 = smallest)
-   $m$ = total number of tests
-   $\alpha$ = target FDR (e.g. 0.05)

As you accept more tests (larger $i$) the threshold relaxes. That is safe because true effects cluster at small p-values and rank early, and the ratio $i/m$ grows slowly, so the procedure spends its error budget gradually rather than all at once.

### Example

Suppose you have 10 tests and want FDR = 0.05:

```{r math-intuition-example}
i <- 1:10
m <- 10
alpha <- 0.05
threshold <- (i / m) * alpha

math_example <- data.frame(
  Rank_i = i,
  `i/m` = i / m,
  `Threshold (i/m)×0.05` = threshold,
  `Interpretation` = c(
    "1st test: very strict",
    "2nd test: slight relaxation",
    "3rd test: more lenient",
    "4th test: continuing to grow",
    "5th test: halfway point",
    "6th test: getting lenient",
    "7th test: fairly relaxed",
    "8th test: quite lenient",
    "9th test: very lenient",
    "10th test: same as unadjusted!"
  )
)

knitr::kable(math_example[1:5, ],
  digits = 4,
  col.names = c("Rank (i)", "i/m", "Threshold", "What this means"),
  caption = "BH threshold calculation (first 5 tests)"
)
```

At rank 1 the threshold is 0.005 (as strict as Bonferroni, 0.05/10); at rank 5 it is 0.025; at rank 10 it is 0.050, the same as no correction at all.

### Why does this control FDR?

The full proof is technical. But an intuition could work: Suppose there are $m_0$ true nulls and $m_1$ true alternatives. Under a true null a p-value is Uniform(0,1), while true alternatives concentrate near 0. Sorting all p-values together pushes the nulls toward the high ranks, where the relaxed threshold still rejects only a controlled fraction of them. Working through the algebra, the expected fraction of false discoveries is at most $\alpha \, (m_0/m) \le \alpha$.

One consequence: the more true effects there are, the better BH does, because the alternatives fill the low ranks and crowd the nulls out.

### More true effects, more protection

```{r math-visualization}
set.seed(444)

# Scenario 1: Few true effects (harder for BH)
n_tests <- 50
n_true_few <- 5
p_true_few <- rbeta(n_true_few, 0.5, 15)
p_null_few <- runif(n_tests - n_true_few, 0, 1)

# Scenario 2: Many true effects (easier for BH)
n_true_many <- 25
p_true_many <- rbeta(n_true_many, 0.5, 15)
p_null_many <- runif(n_tests - n_true_many, 0, 1)

# Create datasets
data_few <- data.frame(
  p_value = c(p_true_few, p_null_few),
  truth = c(rep("True", n_true_few), rep("Null", n_tests - n_true_few)),
  scenario = "10% True Effects"
) %>%
  arrange(p_value) %>%
  mutate(rank = row_number())

data_many <- data.frame(
  p_value = c(p_true_many, p_null_many),
  truth = c(rep("True", n_true_many), rep("Null", n_tests - n_true_many)),
  scenario = "50% True Effects"
) %>%
  arrange(p_value) %>%
  mutate(rank = row_number())

combined_data <- rbind(data_few, data_many) %>%
  mutate(bh_threshold = (rank / n_tests) * 0.05)

ggplot(combined_data, aes(x = rank, y = p_value)) +
  geom_line(aes(y = bh_threshold), color = "blue", linewidth = 1) +
  geom_point(aes(color = truth, shape = truth), size = 2.5, alpha = 0.7) +
  facet_wrap(~scenario, ncol = 2) +
  scale_color_manual(values = c("True" = "#E74C3C", "Null" = "#95A5A6")) +
  scale_shape_manual(values = c("True" = 16, "Null" = 1)) +
  labs(
    title = "More true effects = Better protection against false positives",
    subtitle = "True effects (red) fill the low ranks, pushing nulls (gray) to higher ranks",
    x = "Rank",
    y = "P-value",
    color = "Truth",
    shape = "Truth"
  ) +
  mytheme +
  theme(legend.position = "top")
```

With many true effects (right panel) the reds fill the low ranks and leave fewer openings for nulls, so BH gains power.

## Example: Gene expression analysis

Consider a larger example: 1000 genes, 50 of them truly differentially expressed.

```{r practical-example}
# Simulate gene expression study
set.seed(456)
n_genes <- 1000
n_true_de <- 50 # 50 genes truly differentially expressed

# Simulate p-values
p_values_real <- c(
  rbeta(n_true_de, 0.3, 15), # True DE genes: small p-values
  runif(n_genes - n_true_de, 0, 1) # Non-DE genes: uniform
)

gene_data <- data.frame(
  gene_id = paste0("Gene_", 1:n_genes),
  p_value = p_values_real,
  truth = c(rep("DE", n_true_de), rep("Not DE", n_genes - n_true_de))
) %>%
  arrange(p_value) %>%
  mutate(
    rank = row_number(),
    # Apply different methods
    bh_threshold = (rank / n_genes) * 0.05,
    bh_significant = p_value <= bh_threshold,
    bonf_significant = p_value <= (0.05 / n_genes),
    unadjusted_significant = p_value <= 0.05
  )

# Calculate discoveries and false discoveries for each method
calc_performance <- function(data, sig_col, truth_col = "truth") {
  true_pos <- sum(data[[sig_col]] & data[[truth_col]] == "DE")
  false_pos <- sum(data[[sig_col]] & data[[truth_col]] == "Not DE")
  true_neg <- sum(!data[[sig_col]] & data[[truth_col]] == "Not DE")
  false_neg <- sum(!data[[sig_col]] & data[[truth_col]] == "DE")

  total_discoveries <- true_pos + false_pos
  fdr_actual <- if (total_discoveries > 0) false_pos / total_discoveries else 0

  data.frame(
    True_Positives = true_pos,
    False_Positives = false_pos,
    Total_Discoveries = total_discoveries,
    FDR_Actual = round(fdr_actual, 3),
    Power = round(true_pos / 50, 3) # 50 true DE genes
  )
}

performance <- data.frame(
  Method = c("Unadjusted (α=0.05)", "Bonferroni", "Benjamini-Hochberg"),
  rbind(
    calc_performance(gene_data, "unadjusted_significant"),
    calc_performance(gene_data, "bonf_significant"),
    calc_performance(gene_data, "bh_significant")
  )
)

knitr::kable(performance,
  caption = "Performance Comparison Across Methods",
  digits = 3
)
```

### Discoveries by method

```{r performance-viz}
# Reshape for plotting
perf_long <- performance %>%
  select(Method, True_Positives, False_Positives) %>%
  pivot_longer(
    cols = c(True_Positives, False_Positives),
    names_to = "Type", values_to = "Count"
  )

ggplot(perf_long, aes(x = Method, y = Count, fill = Type)) +
  geom_col(position = "stack", alpha = 0.8) +
  geom_text(aes(label = Count),
    position = position_stack(vjust = 0.5),
    color = "white", fontface = "bold", size = 5
  ) +
  scale_fill_manual(
    values = c(
      "True_Positives" = "#27AE60",
      "False_Positives" = "#E74C3C"
    ),
    labels = c("False Positives", "True Positives")
  ) +
  labs(
    title = "Discoveries by method",
    subtitle = "Green = correct discoveries, Red = false positives",
    x = NULL,
    y = "Number of genes called significant (DE)",
    fill = NULL
  ) +
  mytheme +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```


Unadjusted testing finds many true positives but also many false ones. Bonferroni makes almost no false calls but misses most of the real effects. BH sits between them, keeping good power while holding the FDR down.

The realised false-discovery proportion for BH in this run is
`{r} performance$FDR_Actual[3]`. FDR is controlled *in expectation* (E[V/R] ≤ α),
not on every single run, so this one number can land above or below 0.05; averaged
over many simulations it settles at or under the target.

## Comparing the rejection regions

Overlaying both thresholds shows which genes each method rejects.

```{r method-comparison, fig.width=8, fig.height=5}
comparison_data <- gene_data %>%
  filter(rank <= 200) %>%
  mutate(
    rejection_status = case_when(
      bh_significant & bonf_significant ~ "Both BH & Bonferroni",
      bh_significant & !bonf_significant ~ "BH only",
      !bh_significant & bonf_significant ~ "Bonferroni only",
      TRUE ~ "Neither"
    )
  )

ggplot(comparison_data, aes(x = rank, y = p_value)) +
  geom_line(aes(y = bh_threshold, color = "BH threshold"), linewidth = 1) +
  geom_hline(aes(yintercept = 0.05 / n_genes, color = "Bonferroni threshold"),
    linewidth = 1, linetype = "dashed"
  ) +
  geom_point(aes(shape = truth, fill = rejection_status),
    size = 3, alpha = 0.7, stroke = 1
  ) +
  scale_color_manual(
    name = "Thresholds",
    values = c(
      "BH threshold" = "blue",
      "Bonferroni threshold" = "orange"
    )
  ) +
  scale_shape_manual(values = c("DE" = 21, "Not DE" = 24)) +
  scale_fill_manual(
    name = "Rejection status",
    values = c(
      "Both BH & Bonferroni" = "#27AE60",
      "BH only" = "#3498DB",
      "Bonferroni only" = "#F39C12",
      "Neither" = "white"
    )
  ) +
  labs(
    title = "Compare methods",
    subtitle = "First 200 genes sorted by p-value",
    x = "Rank",
    y = "p-value"
  ) +
  mytheme +
  theme(legend.position = "right") +
  guides(shape = guide_legend(title = "Truth", override.aes = list(fill = "gray")))
```



