
# Challenge Problems

Use everything above to answer three open questions.
You already have `dds`, `counts`, `count_mat`, and `metadata` (with the
XIST-inferred `sex` you computed earlier). Write your own code in the empty chunks.

## Challenge 1: Sex-specific genes in healthy liver (beyond the obvious)

Working **only in the `control` samples**, find genes that differ between male
and female livers. We *expect* X/Y-linked genes (XIST, RPS4Y1, DDX3Y, UTY, ...).

**Question:** Are there autosomal genes that are sex-biased in normal
liver?

Steps to think through:

1. Subset to `control` samples and set the design to `~ sex`.
2. Run DE (male vs female).
3. Split the hits into **sex-chromosome** vs **autosomal** genes -- you'll need a
   gene -> chromosome map (e.g. from `org.Hs.eg.db`, column `"CHR"`, or from the
   GTF you downloaded). Report the autosomal sex-biased genes.


```{r}
#| eval: false
# YOUR CODE HERE
# hint: dds_ctrl <- dds[, dds$stage == "control"]
#       dds_ctrl$sex <- factor(dds_ctrl$sex)
#       design(dds_ctrl) <- ~ sex ; dds_ctrl <- DESeq(dds_ctrl)
#       res_sex <- results(dds_ctrl, contrast = c("sex","male","female"))
#       # then map gene -> chromosome and split X/Y vs autosomal
```

<details><summary>Hint: gene -> chromosome map</summary>

```{r}
#| eval: false
chr_map <- AnnotationDbi::mapIds(org.Hs.eg.db,
             keys = rownames(dds), keytype = "SYMBOL",
             column = "CHR", multiVals = "first")
sex_chr <- names(chr_map)[chr_map %in% c("X", "Y")]
```

</details>

## Challenge 2: Is the sex effect the SAME in disease as in health?

Now use **all** samples. Does any gene's male-vs-female difference
*changes* with disease stagem i.e. a **sex x stage interaction**. A gene that
is sex-biased in controls but not in moderate NAFLD (or vice versa) is an
'interaction' gene.

Steps:

1. Design `~ sex + stage + sex:stage` (drop stages with too few of one sex --
   check `table(metadata$sex, metadata$stage)` first; `control` has only ~10
   samples, so the interaction there will be underpowered -- say so).
2. Use `results()` on the interaction term (see `resultsNames(dds)`), or an LRT:
   `DESeq(dds, test = "LRT", reduced = ~ sex + stage)`.
3. Contrast the sex effect *within* control vs *within* moderate and list genes
   whose sex-difference flips or appears only in one condition.

Read the DESeq2 [vignette](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) to understand how interactions work.

```{r}
#| eval: false
# YOUR CODE HERE
# hint: table(metadata$sex, metadata$stage)   
#       design(dds) <- ~ sex + stage + sex:stage
#       dds <- DESeq(dds, test = "LRT", reduced = ~ sex + stage)
#       res_int <- results(dds)               
```

## Challenge 3: Genes that change *progressively* with severity

`Stage` is only control / early / moderate. But the `group in paper` column has a progressive (clinical) stage of the disease:
**control -> NAFL -> NASH_F0-F1 -> NASH_F2 -> NASH_F3 -> NASH_F4**.
Find genes whose expression **moves monotonically** along this axis (a dose-response / progressive signature): 

Two potential approaches:

- **Ordered / trend test:** encode the ladder as an ordered numeric score
  (control = 0, NAFL = 1, ... NASH_F4 = 5), put it in the design as a continuous covariate, and DE on that coefficient -> genes with a consistent slope.
- **LRT across all levels**, then keep genes whose per-level means are monotone (increasing or decreasing) across the ordered ladder.

Then: are any of your Challenge-1 autosomal sex genes ALSO progressive? Overlap the lists.

```{r}
#| eval: false
# YOUR CODE HERE
# hint (trend test):
# ladder <- c("control","NAFL","NASH_F0-F1","NASH_F2","NASH_F3","NASH_F4")
# metadata$severity <- match(metadata$`group in paper`, ladder) - 1   # 0..5
# dds2 <- DESeqDataSetFromMatrix(count_mat, metadata, ~ severity)
# dds2 <- DESeq(dds2)
# res_trend <- results(dds2, name = "severity")   # slope per unit severity
# # sort by padj; check a few top genes with plotCounts(dds2, gene=..., intgroup="group in paper")
```

Discuss: the trend test assumes a roughly linear response on the log scale. What biology would it miss (e.g. genes that spike only at NASH_F4)? How would you catch those?
