---
title: "Integration: horizontal integration of PBMC3k + PBMC4k, and label transfer"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
date: today
---

# Setup

```{r}
setRepositories(ind = 1:3)
required.packages <- c("Seurat", "BiocManager", "harmony", "glmGamPoi", "remotes", "batchelor")
InstallMissing <- function(packages) {
    for (pkg in packages) {
        if (!require(pkg, character.only = TRUE)) {
            install.packages(pkg, dependencies = TRUE)
            library(pkg, character.only = TRUE)
        }
    }
}

remotes::install_github("satijalab/seurat-wrappers")
```

```{r}
library(Seurat)
library(dplyr)
library(SeuratWrappers)
```

# Load and merge the two datasets

Download the query dataset `pbmc4k` which we will annotate from [here](https://www.dropbox.com/scl/fi/qvy00944y56ja953nf2v9/pbmc4k.rds?rlkey=zit84xgh36pqho4f8lafoo0qn&st=bkopiy8v&dl=0) and the annotated `pbmc3k` dataset from [here](https://www.dropbox.com/scl/fi/tbjbnljsd0iowikfzxeu2/pbmc3k_annotated.rds?rlkey=surqa1nw8th5i7in8st6gmk2j&st=fmcawk75&dl=0). Note that in an ideal world, your reference annotated object
will have more cells than the query.

```{r}
# EDIT THIS
PBMC3k_path <- "~/github/SSGIM2026/single_cell/pbmc3k_annotated.rds"
PBMC4k_path <- "~/github/SSGIM2026/single_cell/pbmc4k.rds"

pbmc3k <- readRDS(PBMC3k_path)
pbmc3k$dataset <- "pbmc3k"
pbmc4k <- readRDS(PBMC4k_path)
pbmc4k$dataset <- "pbmc4k"

# Merge the two datasets
merged <- merge(pbmc3k, y = pbmc4k, add.cell.ids = c("pbmc3k", "pbmc4k"))
```

# Standard workflow on the merged (uncorrected) data

```{r}
merged <- NormalizeData(merged) %>%
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>%
  ScaleData() %>%
  RunPCA(npcs = 30) %>%
  FindNeighbors(dims = 1:20) %>%
  RunUMAP(dims = 1:20)

# Do the datasets merge into each other well?
DimPlot(merged, reduction = "umap", group.by = "dataset")

merged <- FindClusters(merged, resolution = 0.3)
merged$merged_cluster <- Idents(merged)

DimPlot(merged, reduction = "umap", group.by = "merged_cluster", label = TRUE)
DimPlot(merged, reduction = "umap", group.by = c("dataset", "merged_cluster"), label = TRUE)
```

# Integrate: MNN, Harmony, CCA

```{r}
merged <- JoinLayers(merged)
merged[["RNA"]] <- split(merged[["RNA"]], f = merged$dataset)
merged <- NormalizeData(merged) %>% FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA()

integrated <-  IntegrateLayers(object = merged, method = FastMNNIntegration,
                                   orig.reduction = "pca",
                                   new.reduction = "integrated.mnn",
                                          verbose = FALSE)
integrated <- IntegrateLayers(
    object = integrated, method = HarmonyIntegration,
    orig.reduction = "pca", new.reduction = "integrated.harmony",
    verbose = FALSE
)
integrated <- IntegrateLayers(
    object = integrated, method = CCAIntegration,
    orig.reduction = "pca", new.reduction = "integrated.cca",
    verbose = FALSE
)
```

# Compare integrations

```{r}
integrated <- FindNeighbors(integrated, reduction = "integrated.cca", dims = 1:20)
integrated <- FindClusters(integrated, resolution = 0.3, cluster.name = "cca_clusters")
integrated <- RunUMAP(integrated, reduction = "integrated.cca", dims = 1:20,
                      reduction.name = "umap.cca")

DimPlot(integrated, reduction = "umap.cca",
        group.by = c("dataset", "merged_cluster", "cca_clusters"), label = TRUE)

integrated <- FindNeighbors(integrated, reduction = "integrated.harmony", dims = 1:20)
integrated <- FindClusters(integrated, resolution = 0.3, cluster.name = "harmony_clusters")
integrated <- RunUMAP(integrated, reduction = "integrated.harmony", dims = 1:20,
                      reduction.name = "umap.harmony")

DimPlot(integrated, reduction = "umap.harmony",
        group.by = c("dataset", "harmony_clusters", "cca_clusters"), label = TRUE)
```

# Label transfer: annotate pbmc4k from pbmc3k

We can use the same idea to *transfer* annotation from pbmc3k to pbmc4k.

```{r}
pbmc3k <- NormalizeData(pbmc3k) %>%
    FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>%
    ScaleData() %>%
    RunPCA(npcs = 30) %>%
    FindNeighbors(dims = 1:20) %>%
    RunUMAP(dims = 1:20)
pbmc4k <- NormalizeData(pbmc4k) %>%
    FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>%
    ScaleData()  %>%
    RunPCA(npcs = 30) %>%
    FindNeighbors(dims = 1:20) %>%
    RunUMAP(dims = 1:20)

anchors <- FindTransferAnchors(reference = pbmc3k,
                               query = pbmc4k,
                               dims = 1:30,
                               reference.reduction = "pca")
predictions <- TransferData(anchorset = anchors, refdata = pbmc3k$seurat_annotations,
                            dims = 1:30)
pbmc4k <- AddMetaData(pbmc4k, metadata = predictions)
DimPlot(pbmc4k, group.by = "predicted.id")
```

UMAP is only for visualisation -- always check using markers if the results
make sense.

```{r}
markers.B <- FindMarkers(object = pbmc4k, ident.1 = "B", only.pos = TRUE, group.by = "predicted.id")
```
